Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-08-30 Writing/NatureWater/1stRoundRevision/replies/reply_to_reviewers_R1.tex

\textcolor{blue}{
Figure 2 in our revised manuscript shows the difference in irrigated crop yields between the 2nd (medium level of saturated thickness) and 3rd (highest level of saturated thickness) quantiles relative to the 1st quantile (lowest level of saturated thickness). The shaded area denotes the 95$\%$ confidence interval around these response curves - indicating whether the difference in irrigated yield between the different levels of saturated thickness is significant at at least the 5$\%$ level. From Figure 2, it is clearly seen that large water deficits - equivalent to a severe drought event as noted above - lead to irrigated corn and soybean yields that are significantly lower for both corn and soybean when saturated thickness is low compared to when saturated thickness is high. The size of the effect is greatest for corn compared to soybean, and when comparing the third quantile (highest saturated thickness) to the first quantile (lowest saturated thickness). As discussed in our original manuscript and expanded upon in the revised manuscript (page 2 lines 86-108), this reflects the larger water requirements and drought sensitivity of corn along with the non-linearities in the response of well yields to saturated thickness.
}

Your comment: P. 2 states that ``counties with the highest level of
saturated thickness \ldots{} experience \ldots{} productivity close to
rainfed yields''. From the figure it is difficult to tell if the 2nd and
3rd quantile are significantly different for soybean. For corn this does
not seem to be the case.

\textcolor{blue}{
You are right that there are only small differences between yields for both corn and soybean when comparing the second and third quantiles of saturated thickness. As noted in the manuscript, this reflects the non-linearities in the relationships between saturated thickness and well yield, which mean that impacts of aquifer depletion on drought risk are not typically felt until saturated thickness has been lowered sufficiently to reduce well yields to levels that impair ability to meet crop water demands. However, in this part of the text, our intention was not to distinguish the 2nd and 3rd quantiles. Instead, we were seeking to point out that - in years with low levels of water deficit (i.e., wetter or non-drought years) - counties with higher aquifer thickness (2nd and 3rd) have lower yields than counties that have the lowest aquifer thickness (1st quantile). Furthermore, yields of counties with higher levels of saturated thickness are comparable to that of rainfed production - in particular for soybean production. To more clearly highlight these points, we have modified the phrasing of the text (page 2 lines 95-108 in the revised manuscript) to now state:}

\textcolor{blue}{"At low water deficits (below 200mm), Figure 2 illustrates that counties with the highest levels of aquifer thickness (i.e., 3rd quantile in Figure 1) experience statistically lower yields for corn and soybean than counties with the lowest level of aquifer thickness (1st quantile). In case of soybean, irrigated yields for the lowest aquifer thickness category converge to levels comparable to those obtained from rainfed production."}

Your comment: Similarly, the text at the bottom of p.~2 states that
``saturated thickness has a statistically significant impact on per-area
irrigated crop yields in moderate to extreme drought conditions''.
Please clarify how the significance is evaluated here. Which range of
deficits is meant by ``moderate to extreme''?

\textcolor{blue}{We agree that our manuscript would benefit from greater clarity around wording on significance of aquifer depletion impacts on productivity at different levels of drought. In our revised manuscript, we have therefore replace words like "moderate to extreme" with more specific water deficit thresholds/ranges. For the phrase quoted by the reviewer, the text in the revised manuscript (page 2 lines 87-88) has been modified to now state:``in years with water deficit higher than about 750 mm.'' We have also explicitly added the term ``statistically'' when we are talking about statistical significance, as opposed to economic or agronomic significance (i.e. implications) of the results.
}

Your comment: Are the soybean yields considered significantly different
between quantiles? Are only the 1st and 3rd quantile considered
significantly different for corn? - It could even be of interest to
readers which results are not significant here, in contrast to the
greater significance of the effects analyzed in Fig. 2 and Fig. 3.

\textcolor{blue}{As discussed in our responses to earlier comments, Figure 2 and the revised supporting text added to our revised manuscript (page 2 lines 86-94) now provide an in-depth analysis of the statistical significance of yield impacts between different saturated thickness categories. For corn, Figure 2 shows that there are statistically significant differences in yields between the 3rd and 1st quantiles of saturated thickness, and also between the 2nd and 1st quantiles of saturated thickness in the case of extreme water deficits. For soybeans, yields are not significantly different across the three saturated thickness categories - likely due to soybeans having lower water requirements and therefore lower vulnerability to reduced well yield linked to smaller saturated thickness. We have added the following short statement about the insignificance of the impact of saturated thickness for per-area soybean yields on page 2 lines 91-94 of the revised manuscript:}

\textcolor{blue}{"Unlike corn, irrigated soybean yields are not statistically distinguishable among the three aquifer thickness groups. This is likely because soybean is not as water demanding a crop as corn, meaning that even counties with lower saturated thickness can meet the majority of crop water demands as well yield is not a significantly binding constraint on irrigation management decisions."}

\begin{itemize}
\tightlist
\item
  In light of point (1.a), please also consider whether additional
  panels showing tests for specific hypotheses or other visual aids
  could make the Fig. 1 more accessible to readers with regards to the
  statements made. The figure is quite crowded, with many confidence
  bands overlapping. Currently, it takes time for readers to parse where
  the confidence bands for some of the plot lines end. The figure could
  be of greater benefit to readers if it was more accessible.
\end{itemize}

\textcolor{blue}{Thank you for this valuable suggestion. As we have discussed above, we have added a new Figure 2, which provides a more detailed comparison of yield responses for the higher levels (2nd and 3rd quantiles) of saturated thickness compared to the lowest levels (1st quantile) of saturated thickness. We believe this change has made it significantly easier for the readers to understand the impact of saturated thickness on farmers ability to buffer crops against water deficits and drought. For Figure 1, it is important to have all the four curves in the same panel as that would allow the readers to compare across both rainfed and the three irrigated yield categories. Confidence intervals around the curves were removed for visual interpretation, taking on board your comments made about difficulty interpreting the original Figure 1. Instead, we have displayed the estimated curves with confidence intervals individually in separate panels in Figure A1 for corn and A2 for soybean of Appendix A.}

\begin{itemize}
\tightlist
\item
  How were confidence bands determined? Were additional tests performed
  to evaluate their validity? Please consider showing individual data
  points.
\end{itemize}

\textcolor{blue}{Confidence bands are derived from the standard error of the coefficients in the model represented by Equation 1. This is a standard approach for determining confidence intervals from a statistically estimated model, and evaluating alternate measures of defining confidence intervals is beyond the scope of our study. Instead, the role of these confidence intervals is to quantify the uncertainty of the point estimates using standard statistical approaches.}

\begin{itemize}
\tightlist
\item
  P. 2 states that ``for water deficits above around 400mm, yield
  changes become negative for counties with the lowest levels of
  saturated thickness (i.e., 1st quantile in Figure 1). This reflects
  the restrictions that lower saturated thickness and associated
  reductions in well yields place on farmers ability to fully meet crop
  water requirements during periods of more severe or extended
  precipitation deficits''. How valid is this interpretation in view of
  the fact that these counties see a similarly large yield decrease
  below 400mm and that counties with the highest levels of saturated
  thickness even see a much stronger increase in yields with increasing
  deficits?
\end{itemize}

\textcolor{blue}{We believe our interpretation is valid, but accept that this is a hypothesis and could require further testing to substantiate the effect that is beyond the scope of our data. We attribute yield reductions with declining saturated thickness at low levels of water deficit (i.e. wetter conditions) to possible over-irrigation by farmer. Over-irrigation in wetter years has been documented empirically by previous research in the High Plains Aquifer (e.g., Foster et al., 2019; Gibson et al., 2017; Gibson et al., 2019), and excess water is known to negatively affect grain yields through a variety of mechanisms including physical damage to crops, nutrient leaching, delays to planting and harvesting, and other factors (e.g., Li et al., 2019). Reductions in yields in this scenario could also reflect the lower availability of positive heat in wetter years, which would diminish the positive benefits of irrigation and possibly explain some of the convergence to rainfed yields. This alone would not explain the fact that yield declines for low water deficits are greatest for counties with the largest aquifer thickness (and by implication greatest potential to over-irrigate). To reflect these different possible interpretations and their uncertainties, we have modified the text on page 2 lines 95-108 of our revised manuscript to now state:}

\textcolor{blue}{"At low water deficits (below 200mm), Figure 2 illustrates that counties with the highest levels of aquifer thickness (i.e., 3rd quantile in Figure 1) experience statistically lower yields for corn and soybean than counties with the lowest level of aquifer thickness (1st quantile). In case of soybean, irrigated yields for the lowest aquifer thickness category converge to levels comparable to those obtained from rainfed production. These counter-intuitive results reflect a combination of two factors. First, in years with small water deficits there is less need for irrigation and hence irrigated and rainfed yield differences are reduced. Second, years with small or even negative water deficits are associated with wetter conditions during the growing season, which can negatively impact yield if waterlogging occurs leading to greater risks of physical crop damage (e.g., lodging of grains), delays to planting/harvesting, nutrient leaching, and restricted root growth in waterlogged soils (Li et al., 2019). Declines in irrigated yields could be indicative of over-irrigation by farmers, which has been shown to occur in the HPA region in wet years (Foster et al., 2019; Gibson et al., 2017; Gibson et al., 2019), exacerbating these risks. However, the small declines in irrigated yields at low or negative levels of water deficit may also reflect the fact that wetter years are typically cooler, reducing positive heat effects on crop growth and yield formation."}

\textcolor{blue}{New reference added:}

\textcolor{blue}{- Li, Y., Guan, K., Schnitkey, G. D., DeLucia, E., and Peng, B. (2019). Excessive rainfall leads to maize yield loss of a comparable magnitude to extreme drought in the United States. Global Change Biology, 25(7), 2325-2337.}

\begin{itemize}
\tightlist
\item
  Please clarify in the legend or caption that the quantiles indicate
  levels of saturated thickness.
\end{itemize}

\textcolor{blue}{Thank you for this comment. We made changes to all the relevant figures and their captions according to this suggestion.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Figure 2: Please clarify in the text how the significance of
  reductions in the share of irrigated acres is evaluated (e.g.~relative
  to which aquifer depletion level(s)?), and at which level of aquifer
  depletion the reduced irrigation share becomes significant for each
  crop.
\end{enumerate}

\textcolor{blue}{We have added a new figure in Appendix B - Figure B.1 - that presents the estimated difference in the share of irrigated acres at different values of aquifer thickness levels relative to an aquifer thickness of 130 m (equivalent to a very thick aquifer in our third quantile group) including 95\% confidence intervals. This figure shows that the reduction in irrigated acres is statistically significant at the 5\% level for aquifer thickness of below approximately 35 m for corn and 30 m for soybean. Given space constraints imposed for the main body of our manuscript, we decided to place this figure in the appendix as it is supplementary to the main discussion.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  Figure 3:
\end{enumerate}

\begin{itemize}
\tightlist
\item
  Please clarify why this figure has no confidence bands. The text
  describes it as a combination of the effects shown in Fig. 1 and Fig.
  2, which both have confidence bands. Please clarify how significance
  is evaluated.
\end{itemize}

\textcolor{blue}{
Thank you for highlighting this. To maintain consistency with our response to your earlier comments and valuable points about ensuring clarity of our figures, we have decided to not present confidence intervals for this figure as in the case Figure 1 to avoid overcluttering the plot. Instead, we now have added a new Figure 5, which is similar to Figure 2. It compares average (irrigated and rainfed combined) yield for counties with aquifer thickness of 50 m, 90 m, and 130 m against those with 10m. As in Figure 2, the shaded area denotes the 95\% confidence interval and can be used to interpret significance at the 5\% level. Since these numbers are equidistant, they allow us to illustrate non-linearity in the impact of aquifer thickness. Figure 5 shows that for for 50 m, 90 m, and 130 m aquifer thickness levels, yields are significantly higher than for counties with 10 m of aquifer thickness for water deficits above 400 mm for corn and above around 600 mm for soybean. The magnitude of yield gains is greatest for 90 m and 130 m aquifer thickness, but somewhat lower for aquifer thickness of 50 m. These responses indicate the non-linearities in the response of yields to both water deficits and aquifer conditions, which we have also expanded discussion of in the text in the final paragraph of Section 2.3 on page 7 lines 154-164 of our revised manuscript
}

\begin{itemize}
\tightlist
\item
  Relatedly, the text states that deficits beyond 200mm yield
  significant reductions in corn production. 200mm is where the curves
  intersect. Please clarify if this actually is also where significance
  starts, or if significance starts at a higher deficit.
\end{itemize}

\textcolor{blue}{You are correct that the text in the original manuscript referred to the point at which the curves intersect. The threshold level of water deficit at which saturated thickness has statistically significant impact on the response of per-area corn production is approximately 350 mm. We believe this is now more clearly demonstrated in Figure 5, and have also clarifies this point in the revised text linked to this figure (paragraph 2 of section 2.3 on page 6).}

\begin{itemize}
\tightlist
\item
  Please double-check if the y-axis label is correct. The label (``yield
  (tonne/ha)'') is the same as for Fig. 1, which makes it difficult to
  distinguish the different results presented in Fig. 1 and 3.
\end{itemize}

\textcolor{blue}{Yield (tonne/ha) is correct. Figures 4 (previously 3) and 5 (new figure) both report yield. Figures 1 and 2 report "irrigated" and "rainfed" yields separately, while Figures 4 and 5 report "average" yields reflecting the combination of irrigated and rainfed yields conditional on estimated irrigated production shares.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  Discussion:
\end{enumerate}

\begin{itemize}
\tightlist
\item
  P. 6: ``Previous research focused on impacts\ldots{}'' this sentence
  largely repeats a sentence from the introduction, not in the same
  words but in content (and using the same references). Please consider
  removing or modifying the sentence to add new insights for readers.
\end{itemize}

\textcolor{blue}{While we agree that this sentence is similar to one provided in the introduction, we believe it is important to restate this point here in order to provide context when discussing the contribution of our findings to existing literature assessing impacts of drought on agriculture. We therefore have not modified this sentence in our revised manuscript, as the purpose is to reinforce how our analysis differs from the existing state of methods and knowledge in the literature.}

\begin{itemize}
\tightlist
\item
  P. 6: ``\ldots{} demonstrate that the relationship between aquifer
  saturated thickness and drought impacts is highly nonlinear when
  accounting for farmers decisions about irrigated production areas.
  This indicates that sustainable groundwater conservation efforts
  {[}\ldots{]} should be targeted in space and time to avoid depletion
  exceeding critical thresholds \ldots{}'' Is the assumption of
  non-linearity tested? It is difficult to assess the validity of this
  statement based on the figures provided. Please clarify (e.g.~in the
  results) what supports this statement.
\end{itemize}

\textcolor{blue}{We now have statements that clarify this point in the last paragraph of section 2.3 on page 7 of the revised manuscript. Non-linearity in the relationship is visually inferred from examining the shape of the yield and production responses to combinations of weather (represented by water deficit) and aquifer (represented by saturated thickness) conditions. These relationships clearly demonstrate that responses are non-linear - for example increasing curvature of yield responses with higher water deficits, or the variable differences in yields and production when comparing different saturated thickness quantiles. Further formal statistical testing of non-linearity is beyond the scope of our paper given page and word limits, and given that non-linearities can be visually inferred as described above.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{5}
\tightlist
\item
  Methods - data sets:
\end{enumerate}

\begin{itemize}
\tightlist
\item
  P. 7: ``For counties that do not meet this condition, only rainfed
  yield observations are obtained to avoid the possibility that
  irrigated production may be only partly dependent on use of
  groundwater from the HPA.'' - Please consider whether this assumption
  could bias results and how potential biases can be ruled out.
\end{itemize}

\textcolor{blue}{
This will not bias the main conclusions from our the regression analysis, specifically the finding that lower saturated thickness reduces returns to irrigated crop production. Rainfed yields are included as part of the modeling merely to provide a comparison to the magnitude of yield impacts across saturated thickness categories for irrigated production. However, these rainfed yield data do not have any influence on the identification of irrigated yield response to water deficit for the three saturated thickness categories. Moreover, counties with only rainfed production observations are not included in regressions to predict irrigated area production shares, which focus solely on counties with both irrigated and rainfed production data in the same county. We have also added the following text on page 16 lines 352-355 of our revised manuscript to clarify this point:}

\textcolor{blue}{"Note that irrigated area share and average crop yield estimates use  data only from counties that have both irrigated and rainfed production observations. Counties with solely rainfed production observation data (Section 4.1) are excluded from these analyses, and are used solely for estimating impacts of drought on per-area crop yields (Equation 1)."}

\begin{itemize}
\tightlist
\item
  P. 9: Saturated thickness data - the number of monitoring wells would
  be of interest for readers to better judge the data, if available.
\end{itemize}

\textcolor{blue}{Aquifer thickness data were derived from interpolation of spatially dense water level observations from the United States Geological Survey's (USGS) national monitoring network. For the years of our analysis, aquifer thickness maps were generated based on between 14,000 and 20,000 annual point water level observations per year, distributed approximately evenly between counties conditional on variations in county area. While these data are a key input to our analysis, a detailed description of how saturated thickness data were generated is beyond the scope of our paper given the word limits imposed by articles in Nature Water. We refer the reader to  Haacker et al., for further information on the generation of saturated thickness data, which includes details about the number of well observations per year and quality control measures applied in production of saturated thickness maps. We have also added the following text to our revised manuscript (page 13 lines 291-293) to provide basic information on the number and distribution of water level observations used to construct aquifer thickness data:}

\textcolor{blue}{"Annual aquifer thickness maps were derived based on interpolation of 14,000 to 20,000 individual observations of water levels per year, along with detailed maps of both the land surface elevation and bedrock elevation of the HPA."}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{6}
\tightlist
\item
  Methods - regression models:
\end{enumerate}

\begin{itemize}
\tightlist
\item
  P. 9: If the model includes fixed effects, how are the single plot
  lines for each category shown in Fig. 1 derived?
\end{itemize}

### 2023-08-30 Writing/NatureWater/1stRoundRevision/replies/reply_to_reviewers_R1.tex

\textcolor{blue}{
That is because a single county-year was selected in predicting yields. We could pick any county-year without changing any of the conclusions we are drawing. This is because fixed effects are merely a shifter. The relative difference in yield across all the categories is the same no matter what county-year we pick, and the curvature will also be the same. Therefore, presenting lines for all the counties would unnecessarily complicate and crowd the figure without producing any more insights than a single estimated curve for a particular county.
}

\begin{itemize}
\tightlist
\item
  P. 9: Developments over time and independent variables: Please address
  early in the main text which overall trends shape developments in the
  aquifer studied. Please include key references about the case study
  region to inform this, such as Mrad et al.~(2020),
  doi.org/10.1073/pnas.2008383117. In light of this, please discuss the
  appropriateness of the approach. The model accounts for year-specific
  ``shocks'', but not other developments over time (e.g.~lagged effects,
  gradual depletion trends, the introduction of regulation, etc.).
  Relatedly, the analysis only saturated thickness and water deficits to
  explain yields, rather than adding more explanatory variables. The
  advantage of the approach taken here might be the simplicity and
  clarity of assumptions. Readers would benefit from a short explicit
  discussion.
\end{itemize}

\textcolor{blue}{Our analysis considers how aquifer depletion influences the resilience of irrigated crop production to drought risk. As part of this analysis, we explicitly consider trends in aquifer depletion over time through use of time-varying saturated thickness data for counties across the HPA. Consideration of other factors that might cause time variations in the exposure to drought risk are accounted for through the use of year and state fixed effects terms in our analysis. This allows us to isolate the effects of differences in saturated thickness on crop production responses to drought effects, which are the main focus of our paper. Explicit analysis of factors such as introduction of regulations is not feasible with available data, but some of these effects are embedded implicitly in our analysis through use of fixed effects terms. Moreover, our discussion includes detailed text (paragraph 5, section 3) outlining the need for future work to examine how adaptive responses by farmers and regulatory changes may alter relationships between aquifer thickness and drought risks, to which we have also added the suggested reference to support this discussion. Given word constraints and the focus of our paper, we are not in a position to provide a detailed background to the history of groundwater development in the HPA, for which we refer readers to the range of relevant references and supporting texts provided.}

\textcolor{blue}{One important omission from analysis, however, was variations in soil characteristics which might influence relationships between saturated thickness and drought risk at sub-state spatial scales. For example, if an area has sandier soils with lower water holding capacity then the impacts of lower saturated thickness (and therefore lower well yields) may be greater than a county dominated by soils with greater water holding capacities. As previously noted, we cannot include county fixed effects - an alternative way of accounting for soil type variability. This is because the majority of saturated thickness variation in our sample is cross-sectional rather than time-dependent, and hence using county-level fixed effects would remove the majority of saturated thickness variability from our analysis. Therefore, we re-ran our regressions this time including sand percentage, silt percentage, and water holding capacity from SSURGO alongside water deficit and saturated thickness predictors. Inclusion of soil properties did not change the conclusions or significance of our previously reported results.}

\textcolor{blue}{New references added:}

\textcolor{blue}{}

\begin{itemize}
\tightlist
\item
  P. 10: Please provide details on the basis functions in the methods
  section or the supplement.
\end{itemize}

\textcolor{blue}{We have added Appendix D in our revised manuscript, which provides a full overview of the basis functions used in our analysis.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{7}
\tightlist
\item
  Supplementary information: Please also consider providing additional
  information to give readers more clarity on the points outlined above
  in the supplementary information.
\end{enumerate}

\textcolor{blue}{Thank you for this suggestion. We have added a set of Appendices, which provide details about (A) the impact of water deficit on irrigated and average yield with confidence intervals, (B) statistical test of the impact of aquifer depletion on the share of irrigated acres relative to 130m aquifer thickness, (C) explanation of the reason for not including county fixed effects and tests of its potential implications for our results, and (D) how spline basis functions are used to estimate the relationship between two variables in a flexible manner without assuming any functional form. We believe that inclusion of these details have made our paper much stronger, and have referenced the appendices above in response to specific comments.}

\hypertarget{minor-points}{%
\subsection{MINOR POINTS:}\label{minor-points}}

\textcolor{blue}{Thank you for noticing there errors. All of these points have been addressed in the revised manuscript.}

\begin{itemize}
\item[$\boxtimes$]
  M1. Please add line numbers.
\item[$\boxtimes$]
  M2. P. 2: Missing closing parenthesis after ``(i.e., 3rd quantile
  \ldots{}''
\item[$\boxtimes$]
  M3. Fig. 1, 2, 3, and 5: Please add panel tags (``a'', ``b'' \ldots)
  to make it easier for readers to understand the captions.
\item[$\boxtimes$]
  M4. Fig. 3: Closing parenthesis of y-axis title missing
\item[$\boxtimes$]
  M5. P. 9: ``WB'' and ``ST'' should not be in italics, unless ``W'',
  ``B'', ``S'', and ``T'' are separate variables
\item[$\boxtimes$]
  M6. P. 9: ``{[}0 - 33.3\(\%\))j'' - the ``j'' does not seem to belong
  here
\item[$\boxtimes$]
  M7. P. 9: ``with sigma\_j represents the category specific intercept''
  should be ``with sigma\_j representing the category-specific
  intercept'' or something similar
\item[$\boxtimes$]
  M8. P. 9: ``under the generalize additive model framework'' should be
  ``generalized'' and ``modeling'' or something similar
\item[$\boxtimes$]
  M9. P. 9: What is ``mu\_t'' in equation 2? Should ``theta\_t'' in
  equation 1 be ``mu\_t''? Because ``theta\_jk'' is already the spline
  basis function.
\item[$\boxtimes$]
  M10. P. 9: One reference seems to be missing (the text includes
  ``(REF)'').
\end{itemize}

\end{document}
